What We Eval in the Shadows

### 2021-04-09 paper/main.tex
methods

\section{Background and Previous work}

This section provides a short introduction to R and the reflective features of
the language; then looks at the semantics of \eval in R and discusses design
choices; lastly, this paper is put in context of previous work.

@@

While a facility to transform data into code is available in many languages,
some design choices affect the expressive power of \eval. The key choices are
the input format, the environment in which generated code is evaluated and the
reflective operations available to that code. Fig.~\ref{comp} summarizes
several designs.

\begin{figure}[!t]\center\small\begin{tabular}{r@{~}l|l|l|l}\hline
\tiny\sc Language&&\sc\tiny Input&\sc\tiny Scope&\tiny\sc Reflective operations\\\hline
\bf Julia&\cite{julia}     & expression& toplevel         & data\\
\bf Java&\cite{cl}  & bytecode  & classloader       & data\\
\bf JavaScript&\cite{ecoop11}& text      & current, toplevel& data\\
\bf R&\cite{R96}  & expression& programmatic      & data, stack, environment\\\hline
\end{tabular}\caption{Design space}\label{comp}
\end{figure}

@@

The R language exposes a rich reflective interface with four main functions,
\c{eval}, \c{evalq}, \c{eval.parent}, and \c{local}. The first is the most
general with three arguments, an expression, its environment and enclosing
environment. The following simplifies details not relevant to this paper, the
interested reader should consult \citet{hadley}.

\begin{lstlisting}
 eval <- function(e, envir = parent.frame(),
                   encl = if(is.list(envir)) parent.frame() else baseenv()) ...
\end{lstlisting}
Parameter \c e is the \c{expression} to evaluate, \c{envir} is the evaluation
environment, and \c{encl} is used to look up variables not found in \c{envir}.
The default value of latter is either the caller's environment or top level.
\Eval can take several types for \c e, for our purposes on the \c{expression}.
These can be thought of as the abstract syntax trees returned by the parser. An
\c{expression} can be obtained by calling \c{quote} or \c{parse} wit a string:
\begin{lstlisting}
 > quote(a + b)
 # a + b
 > parse(text="a+b")
 # expression(a+b)
\end{lstlisting}
Usually, expressions are obtained from promises using \c{substitute}. Consider a
function called with \c{x = a+b}, then \c f returns an expression object.
\begin{lstlisting}
 > f <- function(x) substitute(x)
 > f(a+b)
 # expression(a+b)
 > substitute(a+b,list(a=1,b=2))
 # 1+2
\end{lstlisting}
The call to \c{substitute(x)} extracts the unevaluated expression from the
promise bound to \c x. Function \c{substitute(e,envir)} takes two arguments, the
second is used as a substitution environment for free variables in \c e.
\begin{lstlisting}
 > environment()
 # <environment: R_GlobalEnv>
\end{lstlisting}
The \c{environment} function returns the current environment. Environments nest,
each has a parent. When creating a new environment with \c{new.env}, the parent
is the current environment. Environment chains can be traversed with
\c{parent.env}, until \c{emptyenv} is reached. The top level environment is
\c{.GlobalEnv}, it has parents that represent the packages that have been
loaded. One can also directly read, modify or create new bindings, given any
environment:
\begin{itemize}
\item \c{envir\$v} and \c{get("v",envir=envir)}: read  \c{v} from \c{envir};
\item \c{envir\$v<-2} and \c{assign("v",2,envir=envir)}: store 2 in \c{v}, if
  not found, \c{v} is created in \c{envir}.
\end{itemize}
\noindent Programmers sometimes use environments as hash maps as they are the
only data type with reference semantics and have a built-in string look up.
Functions \c{parent.frame} and \c{sys.frame} return environments further up the
call stack.
 \begin{lstlisting}
  evalq <- function(e,envir,encl) eval(quote(e),envir,encl)
  eval.parent <- function(e,n) eval(e,parent.frame(n))
  local <- function(e,envir,encl) evalq(e,new.env())
\end{lstlisting}
The three \eval variants can be expressed with \eval. The \c{evalq} form quotes
the argument to prevent it from being evaluated in the current environment. The
\c{eval.parent(e,n)} form evaluates \c{e} in the environment of the {\tt n}-th
caller of this function. Finally, \c{local} evaluates \c{e} in a new environment
to avoid polluting the current one.

@@

\citet{ecoop11} provided the first large-scale study of the runtime behavior of
\eval in JavaScript. They dynamically analyzed a corpus of the 10,000 most
popular websites with an instrumented web browser to gather execution traces.
They show that \eval is pervasive with 82\% of the most popular websites using
it. The reasons for its use include the desire to load code on demand,
deserialization of JSON dataq and lightweight meta-programming to customize web
pages. While many uses were legitimate, just as many were unnecessary and could
be replaced with equivalent and safer code. They categorized inputs to \eval so
as to cover the vast majority of input strings. Restricting themselves to \eval
in which all named variables refer to the global scope, many patterns could be
replaced by more disciplined code~\cite{oopsla12b, moller12}. The work did not
measure code coverage, so the numbers presented are a lower bound on the
possible behaviors. Furthermore, JavaScript usage in 2011 is likely different
from today, e.g. Node.js was not covered by Richards. More details about dynamic
analysis of JavaScript can be found in~\cite{liang}.

@@

\mypara{Base.} The base dataset contains \BasePackages core libraries
performing basic arithmetics, statistics, and operating system related
functionalities. These libraries are bundled with R and are executed
pervasively. Together they contain \BaseEvalCallSites calls to \eval in
\BaseFunsWithEvals functions. Some of these functions provide basic R
functionalities such as package loading and thus there is hardly any R code
which does not invoke one of these \eval call sites.

\mypara{CRAN.} The Comprehensive R Archive
Network\footnote{\url{cran.r-project.org}} is the largest curated repository of
R packages. It hosts \CranAvailablePackagesRnd packages with 6 new one submitted
each day~\cite{Ligges2017}. Packages accepted to CRAN are authored by
experienced developers and abide to a number of well-formedness rules
automatically checked on each commit. Furthermore, each package comes with
sample data. There are three sources of runnable code in a package: \emph{tests,
examples} and \emph{vignettes} -- respectively, unit tests, code snippets from
the documentation, and long-form use-cases written in RMarkdown.
Examples and vignettes can be turned into scripts by extracting the relevant
code. From \CranPackages packages we extracted \CranRunnableScripts scripts that contain
\CranRunnableCode lines of code (\CranRunnableCodeExamplesRnd in examples,
\CranRunnableCodeVignettesRnd in vignettes and \CranRunnableCodeTestsRnd in
tests).

\mypara{Kaggle.} The Kaggle website\footnote{\url{kaggle.com}} is an online
platform for data science. It allows users to submit problems and compete to
solve them. Solutions, called \emph{kernels}, are uploaded as plain scripts or
notebooks. The quality of code on Kaggle is not uniform. Each kernel is
runnable, and thus there is one script per kernel. Input data is provided. We
obtained \KaggleKernels kernels. Since Kaggle is not curated, used SHA-1 hashes
to identify and remove \KaggleDuplicates duplicate entries. The remaining
\KaggleUnique kernels are unique solutions to \KaggleCompetitions competition.
They contain \KaggleCode lines of R code. Only \KaggleWithEvals kernels call
\eval.

@@

\begin{figure}[!h]\hspace{-5mm}
  \includegraphics[width=1.05\linewidth]{pipeline.pdf}
  \caption{Pipeline}\label{fig:pipeline}
\end{figure}

@@

\mypara{Tracer.} The tracer registers callbacks to all \eval functions as well
as a few other ones that allow us to better identify where do the expressions
come from. For example, we taint the results of calls to \c{parse} (turns
string into expression) and \c{match.call} (reflects on current call). We also
capture calls to R API for dynamic code loading (\eg\xspace \c{library},
\c{require} and \c{source}). Next, we subscribe to the events related to a
variable definition and assignment allowing us to record side effects that
happen in environments while evaluating code in \eval. One challenge we
encountered during the analysis was that essentially R only provides source
references for top-level calls in a function body or in the block surrounded
by braces. That is to say, a method whose body would \c{eval(x)} would not have
debug information that we can use to identify that particular \eval. The same
expression surrounded by braces would have a line number and source reference.
This is unfortunately not easily fixed in the R implementation. Instead, we
extended the dynamic analysis tool to attach synthetic source code references
to all \eval call sites by traversing AST.  The tracer is implemented as an R
package in 3.2K C++ and 1.3K of R code. For performance reason, most of the
tracing is done in C++. While in theory the implementation is rather
straightforward, what makes it hard in practice is that (1) some of the core R
functionality such as package loading is implemented using \eval which then
leaks to the tracer in unexpected places, (2) lazy evaluation makes it harder
to analyze function arguments while tracing as prematurely forcing a promise
might have consequences, and finally (3) running large amount of real-world
code triggers exercises a lot of corner cases of highly underspecified R
behavior.

>>>>

\newpage
\section{Background and Previous work}

This section provides a short introduction to R and the reflective features of
the language; then looks at the semantics of \eval in R and discusses design
choices; lastly, this work is put in context.

@@

While a data-to-code facility is available in many languages, some design
choices affect the expressive power of \eval. The key choices are the input
format, the environment in which generated code is evaluated and the reflective
operations available to that code. Fig.~\ref{comp} summarizes designs.

\begin{figure}[!h]\center\small\begin{tabular}{r@{~}l|l|l|l}\hline
\tiny\sc Language&&\sc\tiny Input&\sc\tiny Scope&\tiny\sc Reflective operations\\\hline
\bf Julia&\cite{julia}     & expression& toplevel         & data\\
\bf Java&\cite{cl}  & bytecode  & classloader       & data\\
\bf JavaScript&\cite{ecoop11}& text      & current, toplevel& data\\
\bf R&\cite{R96}  & expression& programmatic      & data, stack, environment\\\hline
\end{tabular}\caption{Design space}\label{comp}
\end{figure}

@@

R exposes a rich reflective interface with functions, \c{eval},
\c{evalq}, \c{eval.parent}, and \c{local}. The following simplifies details not
relevant to this paper, the interested reader should consult \citet{hadley}.

\begin{lstlisting}
 eval <- function(e, envir = parent.frame(),
                   encl = if(is.list(envir)) parent.frame() else baseenv()) ...
\end{lstlisting}
The most general is \eval. Parameter \c e is the \c{expression} to evaluate,
\c{envir} is the evaluation environment, and \c{encl} is used to look up
variables not found in \c{envir}. The default value of latter is either the
caller's environment or top level. \Eval can take several types for \c e, for
our purposes on the \c{expression}. These can be thought of as the abstract
syntax trees returned by the parser. An \c{expression} can be obtained by
calling \c{quote} or \c{parse} with a string:
\begin{lstlisting}
 > quote(a + b)
 # a + b
 > parse(text="a+b")
 # expression(a+b)
\end{lstlisting}
Usually, expressions are obtained from promises using \c{substitute}. Consider a
function called with \c{x = a+b}, then \c f returns an expression object.
\begin{lstlisting}
 > f <- function(x) substitute(x)
 > f(a+b)
 # expression(a+b)
 > substitute(a+b,list(a=1,b=2))
 # 1+2
\end{lstlisting}
The call to \c{substitute(x)} extracts the unevaluated expression from the
promise \c x. Function \c{substitute(e,envir)} takes two arguments, the second
is used to substitute free variables in \c e.
\begin{lstlisting}
 > environment()
 # <environment: R_GlobalEnv>
\end{lstlisting}
The \c{environment} function returns the current environment. Environments nest,
each has a parent. When creating a new environment with \c{new.env}, the parent
is the current environment. Environment chains can be traversed with
\c{parent.env}, until \c{emptyenv} is reached. The top level environment is
\c{.GlobalEnv}, it has parents that represent the packages that have been
loaded. One can also directly read, modify or create new bindings, given any
environment:
\begin{itemize}
\item \c{envir\$v} and \c{get("v",envir=envir)}: read  \c{v} from \c{envir};
\item \c{envir\$v<-2} and \c{assign("v",2,envir=envir)}: store 2 in \c{v}, if
  not found, \c{v} is created in \c{envir}.
\end{itemize}
\noindent Environments are used as hash maps as they have reference semantics
and a built-in string look up. Functions \c{parent.frame} and \c{sys.frame}
return environments further up the call stack.
 \begin{lstlisting}
  evalq <- function(e,envir,encl) eval(quote(e),envir,encl)
  eval.parent <- function(e,n) eval(e,parent.frame(n))
  local <- function(e,envir,encl) evalq(e,new.env())
\end{lstlisting}
The three variants can be expressed with \eval: \c{evalq} quotes its
argument to prevent evaluation in the current environment,
\c{eval.parent(e,n)} evaluates \c{e} in the environment of the {\tt n}-th
caller of this function; finally, \c{local} evaluates \c{e} in a new environment
to avoid polluting the current one.

@@

\citet{ecoop11} provided the first large-scale study of the runtime behavior of
\eval in JavaScript. They dynamically analyzed a corpus of the 10,000 most
popular websites with an instrumented web browser to gather execution traces.
They show that \eval is pervasive with 82\% of the most popular websites using
it. The reasons for its use include the desire to load code on demand,
deserialization of JSON dataq and lightweight meta-programming to customize web
pages. While many uses were legitimate, just as many were unnecessary and could
be replaced with equivalent and safer code. They categorized inputs to \eval so
as to cover the vast majority of input strings. Restricting themselves to \eval
in which all named variables refer to the global scope, many patterns could be
replaced by more disciplined code~\cite{oopsla12b, moller12}. The work did not
measure code coverage, so the numbers presented are a lower bound on possible
behaviors. Furthermore, JavaScript usage in 2011 is likely different from today,
e.g. Node.js was not covered. More details about dynamic analysis of JavaScript
can be found in~\cite{liang}.

@@

\mypara{Base.} The dataset contains \BasePackages libraries performing basic
arithmetics, statistics, and operating system functionalities. These libraries
are bundled with R and executed pervasively. Together they contain
\BaseEvalCallSites calls to \eval in \BaseFunsWithEvals functions. Some of these
functions provide basic functionalities such as package loading and thus there
is hardly any code which does not invoke one of these.

\mypara{CRAN.} The Comprehensive R Archive Network
(\small{\url{cran.r-project.org}}) is the largest curated repository of R
packages. It hosts \CranAvailablePackagesRnd packages with 6 new ones submitted
daily~\cite{Ligges2017}. Packages are authored by experienced developers and
abide by well-formedness rules automatically checked on each commit.
Each package comes with sample data. There are three sources of
runnable code in a package: \emph{tests, examples} and \emph{vignettes} --
respectively, unit tests, code snippets from the documentation, and long-form
use-cases written in RMarkdown. Examples and vignettes can be turned into
scripts by extracting the relevant code. From \CranPackages packages we
extracted \CranRunnableScripts scripts that contain \CranRunnableCode lines of
code (\CranRunnableCodeExamplesRnd in examples, \CranRunnableCodeVignettesRnd in
vignettes and \CranRunnableCodeTestsRnd in tests).

\mypara{Kaggle.} The Kaggle website (\small{\url{kaggle.com}}) is an online
platform for data science. It allows users to submit problems and compete to
solve them. Solutions, called \emph{kernels}, are uploaded as plain scripts or
notebooks. The quality of code on Kaggle is not uniform. Each kernel is
runnable, and thus there is one script per kernel. Input data is provided. We
obtained \KaggleKernels kernels. Since Kaggle is not curated, used SHA-1 hashes
to identify and remove \KaggleDuplicates duplicate entries. The remaining
\KaggleUnique kernels are unique solutions to \KaggleCompetitions competition.
They contain \KaggleCode lines of R code. Only \KaggleWithEvals kernels call
\eval.

@@

\begin{figure}[!h]\hspace{-5mm}
  \includegraphics[width=.95\linewidth]{pipeline.pdf}
  \caption{Pipeline}\label{fig:pipeline}
\end{figure}

@@

The tracer registers callbacks to all \eval functions as well as a few other
ones that allow us to better identify where do the expressions come from. For
example, we taint the results of calls to \c{parse} (turns string into
expression) and \c{match.call} (reflects on current call). We also capture calls
to R API for dynamic code loading (\eg\xspace \c{library}, \c{require} and
\c{source}). Next, we subscribe to the events related to a variable definition
and assignment allowing us to record side effects that happen in environments
while evaluating code in \eval. One challenge we encountered during the analysis
was that essentially R only provides source references for top-level calls in a
function body or in the block surrounded by braces. That is to say, a method
whose body would \c{eval(x)} would not have debug information that we can use to
identify that particular \eval. The same expression surrounded by braces would
have a line number and source reference. This is unfortunately not easily fixed
in the R implementation. Instead, we extended the dynamic analysis tool to
attach synthetic source code references to all \eval call sites by traversing
AST. The tracer is implemented as an R package in 3.2K C++ and 1.3K of R code.
For performance reason, most of the tracing is done in C++. While in theory the
implementation is rather straightforward, what makes it hard in practice is that
(1) some of the core R functionality such as package loading is implemented
using \eval which then leaks to the tracer in unexpected places, (2) lazy
evaluation makes it harder to analyze function arguments while tracing as
prematurely forcing a promise might have consequences, and finally (3) running
large amount of real-world code triggers exercises a lot of corner cases of
highly underspecified R behavior.
